---
title: "Lab 5 assignment - Point pattern analysis"
output:
  html_document:
    df_print: paged
---

```{r, warning=F, error=F, message=F}
rm(list=ls())

require(spatstat)
require(tidyverse)
require(terra)
require(ggplot2)
# Place others as well

# Lcross has i and j arguments. If i is species 1, you are measuring how far species are to species 1. i and j can be species 1 and 2 or both. 

```


# Challenge 1 (3 points)

Create a simulation window on a landscape stretching from 0 to 100 units in both the x and y directions. Now simulate the distribution of an organism on this landscape using a Poisson process with a lambda value of 0.01. Plot this organism's distribution. How many organisms are there on your landscape? What is the expected number of organisms that you should find on this landscape based on your simulation parameters? Why?

```{r}
# Window
SimWindow = owin(xrange=c(0,100), yrange=c(0,100))

# Simulation of organism distribution
set.seed(2024)
OrganismDistribution = rpoispp(lambda = 0.01, win = SimWindow)
summary(OrganismDistribution)

# Plot
plot(OrganismDistribution, main="Distribution of Organisms", xlab="X-coordinate", ylab="Y-coordinate")

# Number of organisms
print(paste("There are", OrganismDistribution$n, "organisms!")) # For Fun
```

As output by my code, there are 109 organisms! Although, the exclamation mark isn't really necessary because based on the simulation parameters, the expected amount would be 100, which is close enough. Given that the simulation window is 100 x 100 units (10,000 units total) and that the lambda value (intensity value) is 0.01, a number near `0.1 x 100 x 100 = 100` should be expected. In other words, our actual value of 109 is 9% higher than what would initially be expected. 

# Challenge 2 (3 points)

Verify that there no linear or quadratic pattern in the distribution of your organisms using fitted Poisson process models. Show the outcome of this analysis and explain your conclusion.

```{r}
# Fit a homogeneous PPM
model_homogeneous <- ppm(OrganismDistribution ~ 1)

# Fit an inhomogeneous PPM with a linear trend
model_linear <- ppm(OrganismDistribution ~ x + y)

# Fit an inhomogeneous PPM with a quadratic trend
model_quadratic <- ppm(OrganismDistribution ~ polynom(x, y, 2))

# Calculate the AIC values of the three models
aic_homogeneous <- AIC(model_homogeneous)
aic_linear <- AIC(model_linear)
aic_quadratic <- AIC(model_quadratic)

# Compare the AIC Values
data.frame(model = c('Homogeneous', 'Linear', 'Quadratic'),
           AIC = c(aic_homogeneous, aic_linear, aic_quadratic))

print(paste("The AIC for the Homogeneous Model is", aic_homogeneous))
print(paste("The AIC for the Linear Model is", aic_linear))
print(paste("The AIC for the Quadratic Model is", aic_quadratic))

```

The model with the lowest AIC value is considered the best. In this scenario, the lowest AIC value belongs to the homogeneous model, indicating neither the linear or quadratic trends improve the model. This means there is no evidence to suggest that there are linear or quadratic patterns in the distribution of points.

# Challenge 3 (14 points)

I have simulated the distributions of 2 critters and provided their x-y coordinates. One of them is a non-territorial predator. The second is a critter that is preyed upon by the predator and exhibits a fear response to the predator's presence. Use the tools you learned in the lab to explore the distributions of these two critters. Identify which one is the predator and which is the prey, and provide code and a thorough justification supporting your conclusion.


```{r}
#Use this code chunk to get you started.

predPreyWin = owin(xrange = c(0, 100), yrange = c(0, 100))

critter1 = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week5/species1.csv') %>% 
  mutate(org = 'critter1')

critter2 = read.csv('https://raw.githubusercontent.com/ValenteJJ/SpatialEcology/main/Week5/species2.csv') %>% 
  mutate(org = 'critter2')
```

Okay, now let's plot the critters.

```{r}

# Convert critter data frames to ppp objects
ppp_critter1 <- ppp(critter1$x, critter1$y, window = predPreyWin)
ppp_critter2 <- ppp(critter2$x, critter2$y, window = predPreyWin)

# Plotting
plot(predPreyWin, main = "Critter Distribution", col = "grey")
plot(ppp_critter1, add = TRUE, pch = 20, col = "blue")
plot(ppp_critter2, add = TRUE, pch = 20, col = "red")

```
Initially, **if I had to guess,** I would say that critter 1 is the prey and critter 2 is the predator. The reason I would guess this is because after plotting the points, it appears that #1 critters behave more similarly to one another. They appear more clumped together, indicating similar repulsion (fear responses) from #2 critter. In contrast, the critter 2 points appear more independent and spread out, indicating they are influenced by the presence of critter 1 points. **Of course, this basic eye-test is not sufficient for the purposes of identify which critter is the predator and which is the prey, so let's dive into some analysis!**

### Let's Check for Ripley's K/L for each individual critter.
Greater aggregation for a set of critters should hint that they are prey. I think of aggregation as the prey huddling up as they try to all get as far as possible from multiple predators. (ADJUST THIS SENTENCE LATER)

```{r}
#Ripley's K without correction for edge effects 
kNone1 = Kest(ppp_critter1, correction='none') # Critter 1
kNone2 = Kest(ppp_critter2, correction='none') # Critter 2
plot(kNone1) # Critter 1
plot(kNone2) # Critter 2
plot(kNone1, .-r~r)
plot(kNone2, .-r~r)

#Ripley's L without correction for edge effects 
lNone1 = Lest(ppp_critter1, correction='none') # Critter 1
lNone2 = Lest(ppp_critter2, correction='none') # Critter 2
plot(lNone1) # Critter 1
plot(lNone2) # Critter 2
plot(lNone1, .-r~r)
plot(lNone2, .-r~r)

#Ripley's K WITH correction for edge effects 
kIso1 = Kest(ppp_critter1, correction='isotropic') # Critter 1
kIso2 = Kest(ppp_critter2, correction='isotropic') # Critter 2
plot(kIso1) # Critter 1
plot(kIso2) # Critter 2
plot(kIso1, .-r~r)
plot(kIso2, .-r~r)

#Ripley's L WITH correction for edge effects 
lIso1 = Lest(ppp_critter1, correction='isotropic') # Critter 1
lIso2 = Lest(ppp_critter2, correction='isotropic') # Critter 2
plot(lIso1) # Critter 1
plot(lIso2) # Critter 2
plot(lIso1, .-r~r)
plot(lIso2, .-r~r)

# Critter 1 Appears more aggregated that expected, especially indicated by final 4 plots


```

Integrate your answer with additional code chunks here.