---
title: "Modeling Indigo Bunting Distributions in Perdido River WMA"
author: "Lucas Parvin"
date: "3-28-2024"
output:
  html_document:
    df_print: paged
---

# **Load Packages**

```{r, warning=F, error=F, message=F}
require(terra)
require(tidyterra)
require(sf)
require(adehabitatHR)
require(adehabitatLT)
require(adehabitatHS)
require(tidyverse)
require(survival)
require(geodata)
require(FedData)
require(tigris)
require(sampling)
require(dismo)
require(predicts)
require(ggnewscale)
require(mgcv)
require(randomForest)
require(maxnet)
require(enmSdmX)
require(gbm)
require(dplyr)
require(caret)
require(ROCR)  
require(pROC)  
require(presenceAbsence) 
require(reshape2)
```


# **Import & Reformat Breeding Bird Data**
```{r, warning=F, error=F, message=F}

visits <- read.csv("Tidy Tables/tidy_BirdLevel2_PointVisits.csv", header = T)
encounters <- read.csv("Tidy Tables/tidy_BirdLevel3_Encounters.csv", header = T)

# Merge 'visits' and 'encounters' data frames
merged_data <- merge(visits, encounters, by = "PointID")

# Selecting data only from Perdido WMA
perdido_birds <- merged_data[merged_data$WMA == "PE", ]

# I used this code to explore NOBO data after the committee meeting
# NOBO <- merged_data[merged_data$Species == "NOBO", ]
# NOBO_Unique_Points <- unique(NOBO$PointID)
# NOBO_Unique_WMAs <- unique(NOBO$WMA)
# NOBO_Unique_Points <- NOBO %>%
#   group_by(WMA) %>%
#   summarise(Unique_Points = n_distinct(PointID))
# print(NOBO_Unique_Points)

# Count the number of each species observed in Perdido
species_counts <- perdido_birds %>%
  group_by(Species) %>%
  summarise(count = n())

# Display the counts
print(species_counts)

# Remove time interval (not relevant for this study)
perdido_birds <- perdido_birds[, -12]

# Assign a presence value of 1 to each row representing INBU, 0 to all others
perdido_birds$INBU <- ifelse(perdido_birds$Species == "INBU", 1, 0)

# Remove all irrelevant covariates
INBU_pres_abs <- perdido_birds[, -c(3,4,5,7,8,11,12,13,14,15),]

# Reduce to unique rows based on all columns
INBU_pres_abs <- unique(INBU_pres_abs)  ## <-- MAY NEED TO CHANGE THIS. SOME SITES HAVE ABSENCE AND PRESENCE
## UP ABOVE, I NEED TO REMOVE ALL ROWS with 0s, if they also have 1s anywhere.

# Here, I ensure each unique point ID is represented only once with a 0 or a 1 in the INBU column. For a 0, the point ID should have previously only been represented with a 0 in the INBU column. For a 1, the previous point ID could have had a single row with an INBU value = 1 or two rows, one with INBU = 0 and on = 1. 
# Step 1: Arrange by PointID and INBU in descending order
# Step 2: Keep only unique rows based on PointID, keeping the first occurrence
INBU_pres_abs <- INBU_pres_abs %>%
  arrange(PointID, desc(INBU)) %>%
  distinct(PointID, .keep_all = TRUE)

# Reduce data set to coordinates and INBU presence-absence values
INBU_pres_abs <- INBU_pres_abs[, -c(1:3)]

```

My code would not push to github when working with all WMAs because the commits were too large. Thus, I decide to just work with Perdido. I chose INBU because they were not ubiquitous, but also well-represented compared to most species.


# **Import & Format Perdido River WMA + Elevation**
```{r, warning=F, error=F, message=F}

# Load Perdido
perdido <- st_read("Perdido_ICP1.shp")

# Transform Perdido to WGS 84 (EPSG:4326)
perdido <- st_transform(perdido, crs = "EPSG:4326")

# Plot Perdido
ggplot(data = perdido) +
  geom_sf() +
  theme_minimal() +
  ggtitle("Perdido")

# Convert 'Perdido' to a SpatVector
perdido_vect <- vect(perdido)
crs_perdido <- crs(perdido_vect)

# Bring in elevation
elevation <- get_ned(perdido_vect, label='Elevation')

# Ensure elevation is projected to correct coordinate system
elevation <- project(elevation, crs_perdido)

# Check to make sure CRS of both 'perdido' and 'elevation' are the same
crs_perdido <- crs(perdido_vect)
crs_elevation <- crs(elevation)
print(crs_perdido)
print(crs_elevation)

# Ensure elevation has the exact same dimensions as Perdido
elevation_clipped <- crop(elevation, perdido_vect)
elevation_clipped <- mask(elevation_clipped, perdido_vect)

# Convert clipped elevation raster to a data frame for plotting
elevation_df <- as.data.frame(elevation_clipped, xy = TRUE)

# Plot Elevation
ggplot() +
  geom_raster(data = elevation_df, aes(x = x, y = y, fill = Layer_1)) +
  scale_fill_viridis_c() + 
  coord_fixed() +
  theme_minimal() +
  labs(title = "Elevation in Perdido WMA", fill = "Elevation") +
  geom_sf(data = perdido, inherit.aes = FALSE, fill = NA, color = "black") # Overlays Perdido boundary

```

# **Derive Slope & Plot**
```{r, warning=F, error=F, message=F}
slope = terrain(elevation, v='slope', neighbors=8, unit='degrees')
slope_clipped <- crop(slope, perdido_vect)
slope_clipped <- mask(slope_clipped, perdido_vect)

slope_df <- as.data.frame(slope_clipped, xy = TRUE)

ggplot(slope_df, aes(x=x, y=y, fill=slope))+
  scale_fill_gradientn(colours = terrain.colors(7))+
  geom_raster()+
  labs(title = "Slope in Perdido WMA") +
  coord_equal()+
  theme_bw()+
  theme(panel.grid=element_blank())
```

# **Derive Aspect & Plot**
```{r, warning=F, error=F, message=F}
aspect = terrain(elevation, v='aspect', neighbors=8, unit='degrees')
aspect_clipped <- crop(aspect, perdido_vect)
aspect_clipped <- mask(aspect_clipped, perdido_vect)

aspect_df <- as.data.frame(aspect_clipped, xy = TRUE)

ggplot(aspect_df, aes(x=x, y=y, fill=aspect))+
  scale_fill_gradientn(colours = terrain.colors(7))+
  geom_raster()+
  labs(title = "Aspect in Perdido WMA") +
  coord_equal()+
  theme_bw()+
  theme(panel.grid=element_blank())
```

# **Bring in NLCD Land Cover Classes, Reclassify, & Plot**
```{r, warning=F, error=F, message=F}
# Bring in NLCD & Ensure raster is aligned with Perdido
nlcd2011 <- get_nlcd(perdido_vect, label='Landscape', year=2008)
nlcd2011 <- project(nlcd2011, crs_perdido )
nlcd2011 <- crop(nlcd2011, perdido_vect)
nlcd2011 <- mask(nlcd2011, perdido_vect)
terra::writeRaster(nlcd2011, 'nlcd.tif', overwrite=TRUE)
nlcd <- rast('nlcd.tif')

# Setting developed land cover class to NA
codes_to_na <- c(21, 22, 23)

# Creating a matrix for reclassification
reclass_matrix <- cbind(codes_to_na, rep(NA, length(codes_to_na)))

# Reclassify the raster
nlcd <- classify(nlcd, reclass_matrix, right=FALSE)

# Assign labels to each NLCD #
land_cover_descriptions <- c(
  "11" = "Open Water",
  "12" = "Perennial Ice/Snow",
  "31" = "Barren Land",
  "41" = "Deciduous Forest",
  "42" = "Evergreen Forest",
  "43" = "Mixed Forest",
  "52" = "Shrub/Scrub",
  "71" = "Grassland/Herbaceous",
  "81" = "Pasture/Hay",
  "82" = "Cultivated Crops",
  "90" = "Woody Wetlands",
  "95" = "Emergent Herbaceous Wetlands"
)

# Convert the raster to a data frame
nlcd_df <- as.data.frame(nlcd, xy = TRUE, na.rm = TRUE)

# Change NLCD #'s to land cover labels
nlcd_df$Class <- factor(land_cover_descriptions[as.character(nlcd_df$Class)])

# Plot
ggplot(nlcd_df, aes(x = x, y = y, fill = Class)) +
  geom_raster() +
  scale_fill_viridis_d() +
  coord_fixed() +
  theme_minimal() +
  labs(title = "NLCD 2008 Land Cover within Perdido Area")

```

# **Load, Format, & Plot Tree Canopy Cover**
```{r, warning=F, error=F, message=F}
treecover <- rast('treecover.tif')

treecover <- project(treecover, crs_perdido )
treecover <- crop(treecover, perdido_vect)
treecover <- mask(treecover, perdido_vect)

treecover_df <- as.data.frame(treecover, xy = TRUE, na.rm = TRUE)

# Plot
ggplot(treecover_df, aes(x = x, y = y, fill = treecover)) +
   geom_raster() +
   coord_fixed() +
   theme_minimal() +
   labs(title = "Tree Canopy Cover in Perdido WMA")


```

# **Ensure All Rasters Are Properly Aligned Before Stacking**
```{r, warning=F, error=F, message=F}

res_standard <- res(elevation)  # Standard resolution
ext_standard <- ext(elevation)  # Standard extent

# Resample and align slope
slope_aligned <- resample(slope, elevation, method="bilinear")
slope_aligned <- crop(slope_aligned, ext_standard)
slope_aligned <- mask(slope_aligned, perdido_vect)

# Resample and align aspect
aspect_aligned <- resample(aspect, elevation, method="bilinear")
aspect_aligned <- crop(aspect_aligned, ext_standard)
aspect_aligned <- mask(aspect_aligned, perdido_vect)

# Resample and align elevation
elevation_aligned <- resample(elevation, elevation, method="bilinear")
elevation_aligned <- crop(elevation_aligned, ext_standard)
elevation_aligned <- mask(elevation_aligned, perdido_vect)

# Resample and align nlcd
nlcd_aligned <- resample(nlcd, elevation_aligned, method="near")
nlcd_aligned <- crop(nlcd_aligned, ext_standard)
nlcd_aligned <- mask(nlcd_aligned, perdido_vect)

# Resample and align tree cover
tree_aligned <- resample(treecover, elevation_aligned, method="near")
tree_aligned <- crop(tree_aligned, ext_standard)
tree_aligned <- mask(tree_aligned, perdido_vect)
```

# **Stack Covariates & Extract Values at Presence-Absence Locations**
```{r, warning=F, error=F, message=F}

covariates = c(elevation_aligned, aspect_aligned, slope_aligned, nlcd_aligned, tree_aligned)
names(covariates) = c('elevation', 'aspect', 'slope', 'NLCD', 'treecover')

coords = cbind(INBU_pres_abs$Longitude, INBU_pres_abs$Latitude)
pres_abs_Covs = terra::extract(covariates, coords)

```


# **Bind INBU Presence-Absence Data with Covariates**
```{r, warning=F, error=F, message=F}
INBU_and_Covs = cbind(pres_abs_Covs, INBU_pres_abs)
INBU_and_Covs$NLCD = as.factor(INBU_and_Covs$NLCD) # Converts NLCD values to factors
```

# **Test for Correlation Between Covariates**
```{r, warning=F, error=F, message=F} 
pairs(covariates, maxpixels=1000)
```

No covariates are significantly correlated, thus I have no need to remove any.

# **Fit GLMs**
```{r, warning=F, error=F, message=F}

glm_intecept = glm(INBU ~ 1, family = 'binomial', data = INBU_and_Covs)
glm_nlcd = glm(INBU ~ NLCD, family = 'binomial', data = INBU_and_Covs)
glm_all = glm(INBU ~ NLCD + treecover + elevation + slope + aspect, family = 'binomial', data = INBU_and_Covs)
glm_interaction = glm(INBU ~ NLCD*elevation, family = 'binomial', data = INBU_and_Covs)
glm_nohab = glm(INBU ~ elevation + slope + aspect, family = 'binomial', data = INBU_and_Covs)
glm_elevtree = glm(INBU ~ treecover + elevation, family = 'binomial', data = INBU_and_Covs)
glm_elevtreeslope = glm(INBU ~ elevation + treecover + slope, family = 'binomial', data = INBU_and_Covs)
```

# **Comparing GLMs**
```{r, warning=F, error=F, message=F}
# Observe the summary of each model
summary(glm_all)
summary(glm_intecept)
summary(glm_nohab)
summary(glm_interaction)
summary(glm_nlcd)
summary(glm_elevtree)
summary(glm_elevtreeslope)

# Define AIC Values
aic_intercept <- AIC(glm_intecept)
aic_nlcd <- AIC(glm_nlcd)
aic_all <- AIC(glm_all)
aic_interaction <- AIC(glm_interaction)
aic_nohab <- AIC(glm_nohab)
aic_elevtree <- AIC(glm_elevtree)
aic_elevtreeslope <- AIC(glm_elevtreeslope)

# Establish minimum AIC (best model)
min_aic <- min(aic_intercept, aic_nlcd, aic_all, aic_interaction, aic_nohab, aic_elevtree, aic_elevtreeslope)

# Define delta AIC values 
delta_aic_intercept <- aic_intercept - min_aic
delta_aic_nlcd <- aic_nlcd - min_aic
delta_aic_all <- aic_all - min_aic
delta_aic_interaction <- aic_interaction - min_aic
delta_aic_nohab <- aic_nohab - min_aic
delta_aic_elevtree <- aic_elevtree - min_aic
delta_aic_elevtreeslope <- aic_elevtreeslope - min_aic

# Create a table that demonstrates model AIC and delta AIC values
summary_table <- data.frame(
  Model = c("glm_intecept", "glm_nlcd", "glm_all", "glm_interaction", "glm_nohab", "glm_elevtree", "glm_elevtreeslope"),
  AIC = c(aic_intercept, aic_nlcd, aic_all, aic_interaction, aic_nohab, aic_elevtree, aic_elevtreeslope),
  Delta_AIC = c(delta_aic_intercept, delta_aic_nlcd, delta_aic_all, delta_aic_interaction, delta_aic_nohab, delta_aic_elevtree, delta_aic_elevtreeslope)
)

# View summary of model AIC and delta AIC values
print(summary_table)

```

**Elevation, tree canopy cover, and slope are consistently the important predictors.** The addition of other covariates and interaction terms worsened AIC values.


# **Predict INBU Presence Across Perdido Using Top GLM**
```{r, warning=F, error=F, message=F}
glmMap = predict(model = glm_elevtreeslope, object = covariates)
glmMap = exp(glmMap)/(1+exp(glmMap))
plot(glmMap, main = "GLM Predicted INBU Presence")

```

For a fair comparison, **all of the remaining models will be fit using the same 3 predictor variables**: elevation, tree canopy cover, and slope.


# **Fit GAM Using & Plot Predicted Presence**
```{r, warning=F, error=F, message=F}
gam_elevtreeslope = gam(INBU ~ s(treecover, k=3) + s(elevation, k=3) + s(slope, k=3), family='binomial', data = INBU_and_Covs, method='ML')
summary(gam_elevtreeslope)
gamMap = predict(covariates, gam_elevtreeslope, type='response')
plot(gamMap, main = "GAM Predicted INBU Presence")
```

# **Fit Boosted Regresssion Tree & Plot Predicted Presence**
```{r, warning=F, error=F, message=F}
set.seed(2024)
boostModel = gbm(INBU ~ elevation + treecover + slope, distribution='bernoulli', n.trees=500, interaction.depth=2, shrinkage=0.1, bag.fraction=0.5, data=INBU_and_Covs)
boostMap = predict(covariates, boostModel, type='response')
boostMap = mask(boostMap, covariates$NLCD) # to clip to Perdido
plot(boostMap, main = "Boosted Regression Predicted INBU Presence")
summary(boostModel)
```

# **Fit Random Forest Model & Plot Predicted Presence**
```{r, warning=F, error=F, message=F}
set.seed(2024)
tuneRF(y = as.factor(INBU_and_Covs$INBU), x=INBU_and_Covs[, c(1,3,5)], stepFactor = 2, ntreeTry = 500)

# OOB is lowest for mtry 1, so mtry 1 is best
rfModel = randomForest(as.factor(INBU) ~ elevation + treecover + slope, data=INBU_and_Covs, mtry=1, ntree=500, na.action = na.omit)
rfMap = predict(covariates, rfModel, type='prob', index=2)
plot(rfMap, main = "Random Forest Predicted INBU Presence")
summary(rfModel)
variable_importance <- importance(rfModel)
print(variable_importance)


```

# **Comparing Parameter Effects on INBU Presence Across Models**

```{r, warning=F, error=F, message=F}
# Elevation
tmp = expand.grid(
  elevation = seq(min(INBU_and_Covs$elevation), max(INBU_and_Covs$elevation), length = 1000),
  treecover = mean(INBU_and_Covs$treecover),
  slope = mean(INBU_and_Covs$slope)
)

elevData = data.frame(
  glm = predict(glm_elevtreeslope, tmp, type = 'response'),
  gam = predict(gam_elevtreeslope, tmp, type = 'response'),
  boost = predict(boostModel, tmp, type = 'response'),
  rf = predict(rfModel, tmp, type = 'prob')[, 2]
) %>% 
  cbind(tmp) %>% 
  dplyr::select(glm:rf, elevation) %>% 
  pivot_longer(cols = glm:rf) %>% 
  mutate(variable = 'elevation')

```

```{r, warning=F, error=F, message=F}
# Tree Cover
tmp = expand.grid(elevation = mean(INBU_and_Covs$elevation),
                  treecover = seq(min(INBU_and_Covs$treecover), max(INBU_and_Covs$treecover), length=1000),
                  slope = mean(INBU_and_Covs$slope))

TreeCoverData = data.frame(
  glm = predict(glm_elevtreeslope, tmp, type = 'response'),
  gam = predict(gam_elevtreeslope, tmp, type = 'response'),
  boost = predict(boostModel, tmp, type = 'response'),
  rf = predict(rfModel, tmp, type = 'prob')[, 2]
) %>% 
  cbind(tmp) %>% 
  dplyr::select(glm:rf, treecover) %>% 
  pivot_longer(cols = glm:rf) %>% 
  mutate(variable = 'treecover')
```

```{r, warning=F, error=F, message=F}
# Slope
tmp = expand.grid(elevation = mean(INBU_and_Covs$elevation),
                  treecover = mean(INBU_and_Covs$treecover),
                  slope = seq(min(INBU_and_Covs$slope), max(INBU_and_Covs$slope), length=1000))

SlopeData = data.frame(
  glm = predict(glm_elevtreeslope, tmp, type = 'response'),
  gam = predict(gam_elevtreeslope, tmp, type = 'response'),
  boost = predict(boostModel, tmp, type = 'response'),
  rf = predict(rfModel, tmp, type = 'prob')[, 2]
) %>% 
  cbind(tmp) %>% 
  dplyr::select(glm:rf, slope) %>% 
  pivot_longer(cols = glm:rf) %>% 
  mutate(variable = 'slope')
```

```{r, warning=F, error=F, message=F}
colnames(elevData)[1] = colnames(TreeCoverData)[1] = colnames(SlopeData)[1] = 'xValue'

tmp = rbind(elevData, TreeCoverData, SlopeData)

ggplot(tmp, aes(x=xValue, y=value, color=name))+
  facet_wrap(~variable, scales='free_x')+
  geom_line()+
  theme_bw()+
  theme(panel.grid=element_blank())


```

**Note that the GAM is hidden by the GLM. Both of these models exhibited the exact same characteristics, outside of their intercept terms.**



# **SDM Validation**

```{r, warning=F, error=F, message=F}

tmp = INBU_and_Covs[, c(1, 3, 5, 8)] ## Isolates pres/abs data and covariates included in models

folds = 5
models = 4

presPoints <- tmp %>%
  filter(INBU == 1)
absPoints <- tmp %>%
  filter(INBU == 0)

# K-fold partitioning
presPoints$kfold = kfold(presPoints, k=folds)
absPoints$kfold = kfold(absPoints, k=folds)
allPoints = rbind(presPoints, absPoints)

# Preparing empty matrices to store data
aucValues = matrix(data = NA, nrow=folds, ncol=models) 
kappaValues = matrix(data = NA, nrow=folds, ncol=models)
sensitivityValues = matrix(data = NA, nrow = folds, ncol = models)
specificityValues = matrix(data = NA, nrow = folds, ncol = models)
tssValues = matrix(data = NA, nrow = folds, ncol = models)

i = 1

for(i in 1:folds){
  training = allPoints[which(allPoints$kfold != i),]
  validation = allPoints[which(allPoints$kfold == i),]
  validation$id = seq(1, nrow(validation), 1)
  
 # Models
  
  glmMod = glm(INBU ~ elevation + treecover + slope, family = 'binomial', data = training)
  gamMod = gam(INBU ~ s(treecover, k=3) + s(elevation, k=3) + s(slope, k=3), family='binomial', data = INBU_and_Covs, method='ML')
  boostMod = gbm(INBU ~ elevation + treecover + slope, distribution='bernoulli', n.trees=100, interaction.depth=2, shrinkage=0.1, bag.fraction=0.5, data=training)
  rfMod = randomForest(as.factor(INBU) ~ elevation + treecover + slope, data=INBU_and_Covs, mtry=1, ntree=500, na.action = na.omit)
  
  # Predictions
  predGlm = predict(glmMod, validation, type='response')
  predGAM = predict(gamMod, validation, type='response')
  predBoost = predict(boostMod, validation, type='response')
  predRf = predict(rfMod, validation, type = 'prob')[,2]
  # rfVal = predict(rfModel, tmp %>% select(canopy:precip), type='prob')[,2], #This line came from Lab 9
           # gamVal = predict(gamModel, tmp %>% select(canopy:precip), type='response'),
  
  # Data Frame

  valDataFrame = data.frame('id' = validation$id,
                            'INBU' = validation$INBU,
                            'predGlm' = predGlm,
                            'predGAM' = predGAM,
                            'predBoost' = predBoost,
                            'predRF' = predRf)
  
  # AUC
  aucValues[i,1] = as.numeric(PresenceAbsence::auc(valDataFrame, which.model=1)[1])
  aucValues[i,2] = as.numeric(PresenceAbsence::auc(valDataFrame, which.model=2)[1])
  aucValues[i,3] = as.numeric(PresenceAbsence::auc(valDataFrame, which.model=3)[1])
  aucValues[i,4] = as.numeric(PresenceAbsence::auc(valDataFrame, which.model=4)[1])
  
  #Find threshold to maximize Kappa
  kappaOpt = PresenceAbsence::optimal.thresholds(valDataFrame, which.model = 1, opt.methods=3)
  kappaValues[i,1] = as.numeric(PresenceAbsence::Kappa(PresenceAbsence::cmx(valDataFrame, which.model = 1, threshold = kappaOpt[[2]]))[1])
  
    kappaOpt = PresenceAbsence::optimal.thresholds(valDataFrame, which.model = 2, opt.methods=3)
  kappaValues[i,2] = as.numeric(PresenceAbsence::Kappa(PresenceAbsence::cmx(valDataFrame, which.model = 2, threshold = kappaOpt[[2]]))[1])
  
    kappaOpt = PresenceAbsence::optimal.thresholds(valDataFrame, which.model = 3, opt.methods=3)
  kappaValues[i,3] = as.numeric(PresenceAbsence::Kappa(PresenceAbsence::cmx(valDataFrame, which.model = 3, threshold = kappaOpt[[2]]))[1])
  
    kappaOpt = PresenceAbsence::optimal.thresholds(valDataFrame, which.model = 4, opt.methods=3)
  kappaValues[i,4] = as.numeric(PresenceAbsence::Kappa(PresenceAbsence::cmx(valDataFrame, which.model = 4, threshold = kappaOpt[[2]]))[1])
}
```


```{r, warning=F, error=F, message=F}

# Calculate mean AUC and Kappa for each model
meanAucValues <- colMeans(aucValues, na.rm = TRUE)
meanKappaValues <- colMeans(kappaValues, na.rm = TRUE)

# Create a data frame to display the results in a table
resultsTable <- data.frame(
  Model = c("GLM", "GAM", "Boosted Regression", "Random Forest"),
  Mean_AUC = meanAucValues,
  Mean_Kappa = meanKappaValues
)

# Print results
print(resultsTable)

# AUC plot
aucPlot <- ggplot(resultsTable, aes(x = Model, y = Mean_AUC, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "black", size = 1) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "black", size = 1) +
  geom_hline(yintercept = 0.9, linetype = "dashed", color = "black", size = 1) +
  annotate("text", x = 1, y = 0.5, label = "Random Chance", vjust = 1.5, hjust = 0.5) +
  annotate("text", x = 1, y = 0.8, label = "Good", vjust = 1.5, hjust = 0.5) +
  annotate("text", x = 1, y = 0.9, label = "Excellent", vjust = 1.5, hjust = 0.5) +
  labs(title = "Mean AUC Values by Model", x = "Model", y = "Mean AUC") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(legend.position = "none")
print(aucPlot)




# Kappa plot
kappaPlot <- ggplot(resultsTable, aes(x = Model, y = Mean_Kappa, fill = Model)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "black", size = 1) +
  geom_hline(yintercept = 0.6, linetype = "dashed", color = "black", size = 1) +
  annotate("text", x = 1, y = 0.8, label = "Almost Perfect Agreement", vjust = -0.5, hjust = 0.5) +
  annotate("text", x = 1, y = 0.6, label = "Substantial Agreement", vjust = 1.5, hjust = 0.5) +
  labs(title = "Mean Kappa Values by Model", x = "Model", y = "Mean Kappa") +
  scale_fill_brewer(palette = "Set1") +
  theme_minimal() +
  theme(legend.position = "none")
print(kappaPlot)



# Plot AUC & Kappa Together
meltedData <- melt(resultsTable, id.vars = "Model", variable.name = "Metric", value.name = "Value")

ggplot(meltedData, aes(x = Model, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = position_dodge(), width = 0.7) +
  theme_minimal() +
  labs(title = "Mean AUC and Kappa Values by Model",
       x = "Model",
       y = "Value",
       fill = "Metric") +
  scale_fill_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



##### ***Code to consider at a later time for my thesis:***

```{r, warning=F, error=F, message=F}
# soil <- get_ssurgo(perdido_vect, label='soil')
# soil <- project(soil, crs_perdido )
# soil <- crop(soil, perdido_vect)
# soil <- mask(soil, perdido_vect)
# terra::writeRaster(soil, 'soil.tif', overwrite=TRUE)
# soil <- rast('soil.tif')
# 
# # Convert the NLCD raster to a dataframe
# soil_df <- as.data.frame(soil, xy = TRUE, na.rm = TRUE)
# 
# # Plotting the NLCD data
# ggplot(soil_df, aes(x = x, y = y, fill = Class)) +
#   geom_raster() +
#   scale_fill_viridis_d(begin = 0.3, end = 0.9, direction = -1, name = "soil") +
#   coord_fixed() +
#   theme_minimal() +
#   labs(title = "soil within Perdido Area")




### Consider TWI (topographic wetness index) ###
### Consider DTW (depth to water) ###



# Maximum Entropy

# pbVect = INBU_and_Covs$INBU
# covs <- INBU_and_Covs %>% dplyr::select(c('elevation', 'treecover', 'slope'))
# 
# maxentModel = maxnet(p = pbVect,
#                      data= covs,
#                      regmult = 1,
#                      classes='lqpht')
# ?maxnet
# 
# plot(maxentModel, type='logistic')
# 
# maxentMap = predictMaxNet(maxentModel, covariates, type='logistic')
# 
# par(mfrow=c(1,1))
# plot(maxentMap)
```