---
title: "Lab 4 Assignment - Landcover"
output:
  html_document:
    df_print: paged
---


```{r, warning=F, error=F, message=F}
# Load Packages test
rm(list=ls())
require(Voss)
require(tidyverse)
require(terra)
require(FedData)
require(sf)
require(tidyterra)
require(landscapemetrics)
require(ggplot2)
```


## Challenge 1 (4 points)

**The landscapemetrics package has functions for calculating 12 patch-level metrics. Calculate all 12 of these for every forest patch in our nlcdSimple raster (using the 8-direction rule). This will result in 12 unique values associated with each forest patch. Use the chart.Correlation() function from the PerformanceAnalytics package to examine the correlation among all 12 of these variables. What patterns do you notice, and what do they tell you about the uniqueness or redundancy of the patch-level metrics you calculated?**

```{r}
studyArea = st_read('/vsicurl/https://github.com/ValenteJJ/SpatialEcology/raw/main/Week4/studyArea.shp')
nlcd = get_nlcd(studyArea, label='AlLandscape', year=2011)

# values(nlcd) = as.character(values(nlcd))

nlcdSimple = nlcd
nlcdSimple[nlcdSimple==11] = 1 #Wet areas are a 1 now
nlcdSimple[nlcdSimple %in% c(21, 22, 23, 24)] = 2 #All developed areas are 2
nlcdSimple[nlcdSimple %in% c(31, 52)] = 3 #Barren land and shrub/scrub are 3
nlcdSimple[nlcdSimple %in% c(41,42,43)] = 4 #All forest types are 4
nlcdSimple[nlcdSimple == 71] = 5 #Grassland is 5
nlcdSimple[nlcdSimple %in% c(81,82)] = 6 #And agriculture is 6

#Give these numbers category names
tmp = data.frame('ID' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))
nlcdSimple = categories(nlcdSimple, value=tmp)

#And plot the new raster
ggplot(nlcdSimple, aes(x=x, y=y, fill=category)) +
  geom_raster()+
  scale_fill_manual(values=c('blue', 'black', 'gray', 'darkolivegreen', 'orange', 'yellow'))
  
  
  
forest = nlcdSimple %>% 
setValues(NA)
  
forest[nlcdSimple ==4] = 1

forestPatchId = patches(forest, directions=8, zeroAsNA=T, allowGaps=F)

plot(forestPatchId)
```

```{r}

# Calculation patch-level charactertics

# Area
patchArea = lsm_p_area(forest, directions=8)
patchArea

# Core area index
patchcai = lsm_p_cai(forest, directions=8)
patchcai

# Circle
patchcircle = lsm_p_circle(forest, directions=8)
patchcircle

# Contiguity index
patchcontig = lsm_p_contig(forest, directions=8)
patchcontig

# Core
patchCore = lsm_p_core(forest, directions=8)
patchCore

# Euclidean Nearest-Neighbor Distance
patchEnn = lsm_p_enn(forest, directions=8)
patchEnn

# Fractal dimension index
patchFrac = lsm_p_frac(forest, directions=8)
patchFrac

# Radius of Gyration
patchGyrate = lsm_p_gyrate(forest, directions=8)
patchGyrate

# Number of core areas
patchNcore = lsm_p_ncore(forest, directions=8)
patchNcore

# Perimeter-Area ratio
patchPara = lsm_p_para(forest, directions=8)
patchPara

# Perimeter
patchPer = lsm_p_perim(forest, directions=8)
patchPer

# Shape
patchShape = lsm_p_shape(forest, directions=8)
patchShape

  
```

```{r}

# Combine metric values into a single data frame
landscape_metrics_df <- data.frame(
  Area = patchArea$value,
  CoreAreaIndex = patchcai$value,
  Circle = patchcircle$value,
  ContiguityIndex = patchcontig$value,
  Core = patchCore$value,
  EuclideanNNDistance = patchEnn$value,
  FractalDimensionIndex = patchFrac$value,
  RadiusOfGyration = patchGyrate$value,
  NumberOfCoreAreas = patchNcore$value,
  PerimeterAreaRatio = patchPara$value,
  Perimeter = patchPer$value,
  Shape = patchShape$value
)

PerformanceAnalytics::chart.Correlation(landscape_metrics_df, histogram=TRUE, pch=19)
  
```

Given that there are 12 patch-level metrics that can be correlated to one another, there are many varied patterns that exist when observing these correlation plots. However, a primary takeaway is that the vast majority of metrics have highly significant correlations with other metrics. Some metrics, such as "Area" and "Core" are completely redundant in this scenario, with a highly significant correlation value equal to 1.00. On the opposite end of the spectrum, "Contiguity" and "Perimeter-area ratio" have a highly signifcant correlation of -0.99, a near perfect inverse. Many other metrics also have highly significant relationships that exhibit more moderate slopes one way or the other. Whereas, others, like "Area" and "Euclidean Nearest-Neighbor Distance" have no signficiant relationship (indicated by lack of astrices). The process of finding all of these values is worthwhile to initially observe their relationships, but using them all in further analysis would be needless for values that strongly predict one another. 


## Challenge 2 (4 points)

**In our lab, we used the 8-direction or "queen" rule to delineate patches. Using the nlcdSimple raster we created, explore the differences in patch characteristics if you were to use the 4-direction or "rook" rule for delineating patches. Calculate the following class-level metrics for forest cover for both the queen and rook patch delineation rules: number of patches, mean patch size, standard deviation in patch size, nearest-neighbor distance, and total edge length. What patterns do you notice? When might it be appropriate to use one rule vs. the other?**

```{r}

#Queen (8)

# Number of Patches
NumPatch8 = lsm_c_np(forest, directions=8)
NumPatch8

# Mean Patch Size
MeanPatch8 = lsm_c_area_mn(forest, directions=8)
MeanPatch8

# Standard Deviation in Patch Size
SDPatch8 = lsm_c_area_sd(forest, directions=8)
SDPatch8

# Mean Nearest-neigbor Distance
NNDistance8 = lsm_c_enn_mn(forest, directions=8)
NNDistance8


# --> Create New Forest Layer for Computing Total Edge <--

forest0 = forest
forest0[is.na(forest)]=0

# Total Edge Length
TEdge8 = lsm_c_te(forest0, directions=8)
TEdge8



#Rook (4)

# Number of Patches
NumPatch4 = lsm_c_np(forest, directions=4)
NumPatch4

# Mean Patch Size
MeanPatch4 = lsm_c_area_mn(forest, directions=4)
MeanPatch4

# Standard Deviation in Patch Size
SDPatch4 = lsm_c_area_sd(forest, directions=4)
SDPatch4

# Mean Nearest-neigbor Distance
NNDistance4 = lsm_c_enn_mn(forest, directions=4)
NNDistance4


# Total Edge Length
TEdge4 = lsm_c_te(forest0, directions=4)
TEdge4


# Place your R code for this challenge here.
```

As is to be expected, under Queen connectivity rules (8 directions), there are fewer patches, but on average they are larger. This contrasts with a higher number of patches of smaller average sizes under Rook connectivity rules (4 directions). Standard deviation in patch size was higher under Queen connectivity than Rook connectivity, likely a result of the larger patches contrasting more greatly with the small specks of patches on the landscape. Nearest Neighbor patches were closer (or of a smaller value) under the Rook connectivty, likely a result of the increased number of patches overall, increasing the likelihood of a nearby patch. Lastly, total edge length was the exact same under both connectivity levels according to the package run. This seems to make sense, given that regardless of whether or not you consider an area 1 or more plots based on a shared vertex (at a diagonal point), the perimeter of the area bordering non-habitat should remain the same. 

I feel that the biology for a species or group of species being studied should be considered when deciding between using the the Queen or Rook connectivity rules. For species with high levels of mobility, using Queen connectivty rules to define patches may be more appropriate, since the units of habitat, which may be connected diagonally, are still easy to move back and forth between. In contrast, for species that exhibit lower mobility, using the Rook connectivity rules may be more appropriate, given that a very small corridor/gap leading to another larger chunk of habitat may not be realistically used or accessible. 



## Challenge 3 (4 points)


**Using the same zoomed-in study area that we used in the lab, download NLCD raster data for the years 2001 and 2019 and simplify these rasters into 6 landcover categories (wet, developed, barren, forest, grassland, and agriculture). Plot these two rasters. What are some of the initial changes you notice between 2001 and 2019?**

```{r}
# Pull NLCD Data For Both Years
nlcd2001 = get_nlcd(studyArea, label='AlLandscape', year=2001)
nlcd2019 = get_nlcd(studyArea, label='AlLandscape', year=2019)

# Simplify 2001
nlcd2001 = nlcd2001
nlcd2001[nlcd2001==11] = 1 #Wet areas are a 1 now
nlcd2001[nlcd2001 %in% c(21, 22, 23, 24)] = 2 #All developed areas are 2
nlcd2001[nlcd2001 %in% c(31, 52)] = 3 #Barren land and shrub/scrub are 3
nlcd2001[nlcd2001 %in% c(41,42,43)] = 4 #All forest types are 4
nlcd2001[nlcd2001 == 71] = 5 #Grassland is 5
nlcd2001[nlcd2001 %in% c(81,82)] = 6 #And agriculture is 6

#Give these numbers category names
tmp = data.frame('ID' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))
nlcd2001 = categories(nlcd2001, value=tmp)


# Simplify 2019
nlcd2019 = nlcd2019
nlcd2019[nlcd2019==11] = 1 # Wet areas are a 1 now
nlcd2019[nlcd2019 %in% c(21, 22, 23, 24)] = 2 # All developed areas are 2
nlcd2019[nlcd2019 %in% c(31, 52)] = 3 # Barren land and shrub/scrub are 3
nlcd2019[nlcd2019 %in% c(41,42,43)] = 4 # All forest types are 4
nlcd2019[nlcd2019 == 71] = 5 # Grassland is 5
nlcd2019[nlcd2019 %in% c(81,82)] = 6 # And agriculture is 6

# Give these numbers category names
tmp = data.frame('ID' = c(1, 2, 3, 4, 5, 6),
                 'category' = c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture'))
nlcd2019 = categories(nlcd2019, value=tmp) # This line may need adjustment based on actual function to assign categories



# Plot 2001 and 2019 NLCD data

# 2001
ggplot(nlcd2001, aes(x=x, y=y, fill=category)) +
  geom_raster()+
  scale_fill_manual(values=c('blue', 'black', 'gray', 'darkolivegreen', 'orange', 'yellow'))

# 2019
ggplot(nlcd2019, aes(x=x, y=y, fill=category)) +
  geom_raster() +
  scale_fill_manual(values=c('blue', 'black', 'gray', 'darkolivegreen', 'orange', 'yellow'))



```

Between 2001 and 2019, a significant amount of forest seems to have been lost. Forest looks to have been replaced by grassland, agriculture, and open areas, presumably all a function of agricultural expansion. 

**Quantify this at the class level by calculating and reporting the changes in (1) the total amount of each land cover type (2) mean patch size for each land cover type, and (3) mean nearest neighbor distance for each cover type between the years 2011 and 2019. Give a short description of how you interpret the changes in these values.**

```{r}
# Total Amount of Each Land Cover Type

# 2001
# Calculation
TotArea2001 = lsm_c_ca(nlcd2001, directions=8)

# Change numeric classes to category names
class_labels <- c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture')
TotArea2001$class <- factor(TotArea2001$class, levels = 1:6, labels = class_labels)

TotArea2001

# 2019
# Calculation
TotArea2019 = lsm_c_ca(nlcd2019, directions=8)

# Change numeric classes to category names
class_labels <- c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture')
TotArea2019$class <- factor(TotArea2019$class, levels = 1:6, labels = class_labels)

TotArea2019

# Difference
Area_Change <- TotArea2019$value - TotArea2001$value
names(Area_Change) <- class_labels
Area_Change
#########################################################################################

# Mean Patch Size of Each Land Cover Type

# 2001
# Calculation
MeanPatch2001 = lsm_c_area_mn(nlcd2001, directions=8)

# Change numeric classes to category names
class_labels <- c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture')
MeanPatch2001$class <- factor(MeanPatch2001$class, levels = 1:6, labels = class_labels)

MeanPatch2001

# 2019
# Calculation
MeanPatch2019 = lsm_c_area_mn(nlcd2019, directions=8)

# Change numeric classes to category names
class_labels <- c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture')
MeanPatch2019$class <- factor(MeanPatch2019$class, levels = 1:6, labels = class_labels)

MeanPatch2019

# Difference
MeanPatch_Change <- MeanPatch2019$value - MeanPatch2001$value
names(MeanPatch_Change) <- class_labels
MeanPatch_Change
#########################################################################################

# Mean Nearest Neighbor Distance of Each Land Cover Type

# 2001
# Calculation
MeanNND2001 = lsm_c_enn_mn(nlcd2001, directions=8)

# Change numeric classes to category names
class_labels <- c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture')
MeanNND2001$class <- factor(MeanNND2001$class, levels = 1:6, labels = class_labels)

MeanNND2001

# 2019
# Calculation
MeanNND2019 = lsm_c_enn_mn(nlcd2019, directions=8)

# Change numeric classes to category names
class_labels <- c('wetland', 'developed', 'open', 'forest', 'grassland', 'agriculture')
MeanNND2019$class <- factor(MeanNND2019$class, levels = 1:6, labels = class_labels)

MeanNND2019

# Difference
MeanNND_Change <- MeanNND2019$value - MeanNND2001$value
names(MeanNND_Change) <- class_labels
MeanNND_Change
```

As demonstrated by the labeled change vectors, the following metrics have been changed in the following ways between 2001 and 2019:
(1) Total amount of each land cover type 
- Wetland: No Change
- Developed: Gain of 1.35 hectares
- Open: Gain of 159.39 hectares
- Forest: Loss of 413.46 hectares
- Grassland: Gain of 125.64 hectares
- Agriculture: Gain of 127.08 hectares
(2) Mean patch size for each land cover type
- Wetland: No Change
- Developed: Loss of 0.18 hectares
- Open: Gain of 2.53 hectares
- Forest: Loss of 31.16 hectares
- Grassland: Gain of 1.77 hectares
- Agriculture: Gain of 4.78 hectares
(3) Mean nearest neighbor distance for each cover type
- Wetland: No Change
- Developed: Loss of 6.42 meters
- Open: Loss of 197.80 meters
- Forest: Loss of 4.26 meters
- Grassland: Loss of 64.44 meters
- Agriculture: Loss of 3.12 meters


**Quantify these changes at the landscape level by calculating and reporting on changes in the (1) Shannon diversity and (2) Shannon evenness of the landscapes at the different time points. Give a short description of how you interpret the changes in these values.**

```{r}

# Shannon Diversity

# 2001
# Calculation
SDiversity2001 = lsm_l_shdi(nlcd2001)
SDiversity2001

# 2019
# Calculation
SDiversity2019 = lsm_l_shdi(nlcd2019)
SDiversity2019

# Difference
SDiversity_Change <- SDiversity2019$value - SDiversity2001$value
SDiversity_Change


# Shannon Evenness

# 2001
# Calculation
SEvenness2001 = lsm_l_shei(nlcd2001)
SEvenness2001

# 2019
# Calculation
SEvenness2019 = lsm_l_shei(nlcd2019)
SEvenness2019

# Difference
SEvenness_Change <- SEvenness2019$value - SEvenness2001$value
SEvenness_Change


```

As demonstrated in the code above, from 2011 to 2019, the Shannon's Diversity Index for the landscape classes increased by 0.3930831 and Shannon's evenness index increased by 0.2193839. These results demonstrate that landscape has become more diversified, in that there is more of an even representation of each land cover type. Given the same 6 land cover types are observed in both years, there is no gain or loss of an entire class. From a wildlife perspective, this diversification in land cover type is not inherently beneficial because some of the cover types (i.e agriculture, developed, etc.) are generally poor habitat for many species. Moreover, fragmentation of land cover types can have mixed effects on wildlife communities, dependent on species-specific adaptations. 



## Challenge 4 (4 points)

**Use the voss2d() function to simulate a surface where g = 7 and H = 0.5. From that, create 9 'landscapes' with 10%, 20%, 30%, ..., 90% threshold values. The '1' values here can represent anything your imagination comes up with. It could be forest cover, cover of some other land cover type, bodies of water, temperatures above a threshold, etc. I suggest you set the seed for your simulation value so that you get the same outcome each time you run the code. Plot these landscapes and comment on what patterns you see changing as the value increases from a 10% cover to 90% cover.**

```{r}
set.seed(23)

vossModel = voss2d(g=7, H=0.5)
vossModel = rast(vossModel$z)

plot(vossModel)

# 10% Forest cover
threshold10 = quantile(as.matrix(vossModel), prob=0.1)
voss10 = ifel(vossModel > threshold10, 0, 1)
plot(voss10)

# 20% Forest cover
threshold20 = quantile(as.matrix(vossModel), prob=0.2)
voss20 = ifel(vossModel > threshold20, 0, 1)
plot(voss20)

# 30% Forest cover
threshold30 = quantile(as.matrix(vossModel), prob=0.3)
voss30 = ifel(vossModel > threshold30, 0, 1)
plot(voss30)

# 40% Forest cover
threshold40 = quantile(as.matrix(vossModel), prob=0.4)
voss40 = ifel(vossModel > threshold40, 0, 1)
plot(voss40)

# 50% Forest cover
threshold50 = quantile(as.matrix(vossModel), prob=0.5)
voss50 = ifel(vossModel > threshold50, 0, 1)
plot(voss50)

# 60% Forest cover
threshold60 = quantile(as.matrix(vossModel), prob=0.6)
voss60 = ifel(vossModel > threshold60, 0, 1)
plot(voss60)

# 70% Forest cover
threshold70 = quantile(as.matrix(vossModel), prob=0.7)
voss70 = ifel(vossModel > threshold70, 0, 1)
plot(voss70)

# 80% Forest cover
threshold80 = quantile(as.matrix(vossModel), prob=0.8)
voss80 = ifel(vossModel > threshold80, 0, 1)
plot(voss80)

# 90% Forest cover
threshold90 = quantile(as.matrix(vossModel), prob=0.9)
voss90 = ifel(vossModel > threshold90, 0, 1)
plot(voss90)

```
As forest cover increases, it mostly expands from the previously established forest cover, rather than popping up randomly in the middle of non-forest patches. Thus, as the forest cover expands, it seems to decrease the total number of forest patches as they become connected, while increasing average patch size. Furthermore, as forest cover increases, intuitively, distances to the nearest neighbor patches seem to decrease, given there is more cover overall to be close to.

**Identify 3 class-level or landscape-level metrics that help you capture the changes you are observing. Calculate those metrics for each of the 9 landscapes and plot them (x-axis is threshold value and y-axis is calculated metric). Briefly describe why you chose these 3 metrics and how they change with increasing cover.**

```{r}

# Number of Patches
NumPatchVoss10 = lsm_c_np(voss10, directions=8)
NumPatchVoss10 <- NumPatchVoss10$value[2]
NumPatchVoss20 = lsm_c_np(voss20, directions=8)
NumPatchVoss20 <- NumPatchVoss20$value[2]
NumPatchVoss30 = lsm_c_np(voss30, directions=8)
NumPatchVoss30 <- NumPatchVoss30$value[2]
NumPatchVoss40 = lsm_c_np(voss40, directions=8)
NumPatchVoss40 <- NumPatchVoss40$value[2]
NumPatchVoss50 = lsm_c_np(voss50, directions=8)
NumPatchVoss50 <- NumPatchVoss50$value[2]
NumPatchVoss60 = lsm_c_np(voss60, directions=8)
NumPatchVoss60 <- NumPatchVoss60$value[2]
NumPatchVoss70 = lsm_c_np(voss70, directions=8)
NumPatchVoss70 <- NumPatchVoss70$value[2]
NumPatchVoss80 = lsm_c_np(voss80, directions=8)
NumPatchVoss80 <- NumPatchVoss80$value[2]
NumPatchVoss90 = lsm_c_np(voss90, directions=8)
NumPatchVoss90 <- NumPatchVoss90$value[2]


# Assuming NumPatchVossXX$value extracts the number of patches from the results
data <- data.frame(
  Threshold = c(0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9),
  NumPatches = c(NumPatchVoss10, NumPatchVoss20, NumPatchVoss30,
                 NumPatchVoss40, NumPatchVoss50, NumPatchVoss60,
                 NumPatchVoss70, NumPatchVoss80, NumPatchVoss90)
)

library(ggplot2)

ggplot(data, aes(x = Threshold, y = NumPatches)) +
  geom_point() +  # Add points
  geom_line() +   # Connect points with lines
  theme_minimal() +
  labs(title = "Number of Patches vs Threshold",
       x = "Threshold",
       y = "Number of Patches")




# Mean Patch Size
MeanPatch4 = lsm_c_area_mn(forest, directions=8)
MeanPatch4

# Mean Nearest-neigbor Distance
NNDistance4 = lsm_c_enn_mn(forest, directions=8)
NNDistance4


```

Place your text here.

## Challenge 5 (4 points)

**Use the voss2d() function to simulate 9 surfaces. In each one, g should be 7, but vary the value of H from 0.1 to 0.9. Create a unique landscape from each of these with a threshold value of 30% cover. Again, the 1 values can represent anything you desire, and again I suggest you set the seed for your simulation value so that you get the same outcome each time you run the code. Plot these landscapes and comment on what patterns you see changing as the H value increases from 0.1 to 0.9.**

```{r}
set.seed(23)

# Place your R code here.

```

Place your text here.

**Identify 3 class-level or landscape-level metrics that help you capture the changes you are observing. THESE MUST BE DIFFERENT THAN THOSE METRICS YOU USED IN CHALLENGE 2. Calculate those metrics for each of the 9 landscapes and plot them (x-axis is H-value and y-axis is calculated metric). Briefly describe why you chose these 3 metrics and how they change with increasing cover.**

```{r}
# Place your R code here.
```

Place your text here.

