---
title: "Publishable, reproducible map of sample sites"
output: html_notebook
---

# Load Useful Packages
```{r}
require(tidyverse)
require(sf)
require(tigris)
require(geodata)
require(ggplot2)
require(terra)
require(colorspace)
require(tidyterra)
require(ggnewscale)
require(cowplot)
```

# Plot Autauga WMA Shapefile
```{r}

# URL of the zipped shapefile on GitHub
url <- "https://github.com/lucasparvin/Spatial_Ecology/raw/main/Week_2_Lab/Assignment/AutaugaWMA.zip"

# Download the file
download.file(url, destfile = "AutaugaWMA.zip", mode = "wb")

# Unzip the file
unzip("AutaugaWMA.zip", exdir = "AutaugaWMA")


# Load the shapefile
autauga_wma <- vect("AutaugaWMA/AutaugaWMA/AutaugaWMA.shp")




# URL of the zipped raster data on GitHub
url <- "https://github.com/lucasparvin/Spatial_Ecology/raw/main/Week_2_Lab/Assignment/autaugalc.zip"

# Download the file
download.file(url, destfile = "autaugalc.zip", mode = "wb")

# Unzip the file
unzip("autaugalc.zip", exdir = "autaugalc")

# Load the raster data (adjust the path if needed)
landcover_raster <- rast("autaugalc/autaugalc") # Replace 'your_raster_folder' with the actual folder name



# Load Alabama shape
alabama <- states() %>% 
  filter(NAME == 'Alabama')

# Extract the CRS from the Alabama shape in WKT format
crs_alabama_wkt <- st_crs(alabama)$wkt

# Transform the CRS of Autauga WMA
autauga_wma_transformed <- terra::project(autauga_wma, crs_alabama_wkt)



```
# Zoomed Out Plot Without extent rectangle
```{r}
library(tidyverse)
library(sf)
library(terra)
library(tigris)

# Load and transform Autauga WMA shapefile
autauga_wma <- vect("AutaugaWMA/AutaugaWMA/AutaugaWMA.shp")
crs_alabama <- crs(states() %>% filter(NAME == 'Alabama'))
autauga_wma_transformed <- terra::project(autauga_wma, crs_alabama)
autauga_wma_sf <- st_as_sf(autauga_wma_transformed)

# Load and transform landcover raster
landcover_raster <- rast("autaugalc/autaugalc")
landcover_raster_transformed <- project(landcover_raster, crs_alabama)
raster_df <- as.data.frame(landcover_raster_transformed, xy = TRUE)
colnames(raster_df) <- c("x", "y", "value")

# Load Alabama shape as sf object
alabama_sf <- states() %>% 
  filter(NAME == 'Alabama') %>% 
  st_as_sf()



# Define plot
ggplot() +
  # Plot the raster
  geom_raster(data = raster_df, aes(x = x, y = y, fill = value)) +
  scale_fill_viridis_c() +  # Color scale for raster
  
  # Overlay Alabama shape
  geom_sf(data = alabama_sf, fill = NA, color = "black", size = 0.5) +
  
  # Overlay Autauga WMA
  geom_sf(data = autauga_wma_sf, fill = NA, color = "red", size = 0.5) +
  
  # Define plot labels and limits
  ggtitle("Alabama with Landcover and Autauga WMA") +
  coord_sf() +  # Adjust coordinate system
  
  # Adjust plot theme
  theme_minimal()



```

#Adding extent rectangle
```{r}
# Calculate the bounding box (extent) of autauga_wma_sf
wma_bbox <- st_bbox(autauga_wma_sf)


ggplot() +
  # Plot the raster
  geom_raster(data = raster_df, aes(x = x, y = y, fill = value)) +
  scale_fill_viridis_c() +  # Color scale for raster
  
  # Overlay Alabama shape
  geom_sf(data = alabama_sf, fill = NA, color = "black", linewidth = 0.5) +
  
  # Overlay Autauga WMA
  geom_sf(data = autauga_wma_sf, fill = NA, color = "red", linewidth = 0.5) +
  
  # Add an extent rectangle for Autauga WMA
  geom_rect(aes(xmin = wma_bbox$xmin, xmax = wma_bbox$xmax,
                ymin = wma_bbox$ymin, ymax = wma_bbox$ymax),
            color = "blue", fill = NA, linetype = "dashed", linewidth = 0.5) +
  
  # Define plot labels and limits
  ggtitle("Alabama with Landcover and Autauga WMA") +
  coord_sf() +  # Adjust coordinate system
  
  # Adjust plot theme
  theme_minimal()



```

# Zooming to extent

```{r}
# Calculate the bounding box (extent) of autauga_wma_sf
wma_bbox <- st_bbox(autauga_wma_sf)

ggplot() +
  # Plot the raster with zoomed extent
  geom_raster(data = raster_df, aes(x = x, y = y, fill = value)) +
  scale_fill_viridis_c() +
  
  # Overlay Alabama shape
  geom_sf(data = alabama_sf, fill = NA, color = "black", linewidth = 0.5) +
  
  # Overlay Autauga WMA
  geom_sf(data = autauga_wma_sf, fill = NA, color = "red", linewidth = 0.5) +
  
  # Add an extent rectangle for Autauga WMA (optional, as we're zooming in)
  geom_rect(aes(xmin = wma_bbox$xmin, xmax = wma_bbox$xmax,
                ymin = wma_bbox$ymin, ymax = wma_bbox$ymax),
            color = "blue", fill = NA, linetype = "dashed", linewidth = 0.5) +
  
  # Define plot labels
  ggtitle("Zoomed into Autauga WMA with Landcover") +
  
  # Zoom into the WMA extent
  coord_sf(xlim = c(wma_bbox$xmin, wma_bbox$xmax), 
           ylim = c(wma_bbox$ymin, wma_bbox$ymax), expand = FALSE) +
  
 # Adjust plot theme
  theme_minimal()


```



# Reclassifying raster values

```{r}

# Landcover values
landcover_values <- c(11, 21, 22, 23, 31, 41, 42, 43, 52, 72, 81, 82, 90, 95)

# Corresponding colors
landcover_colors <- c("11" = "blue", "21" = "pink", "22" = "red", "23" = "red",
                      "31" = "brown", "41" = "lightgreen", "42" = "darkgreen", 
                      "43" = "green", "52" = "tan", "72" = "yellow", 
                      "81" = "orange", "82" = "brown", "90" = "lightblue", 
                      "95" = "purple")

# Landcover labels
landcover_labels <- c("11" = "Open Water", "21" = "Developed Open, Open Space", "22" = "Developed, Low Intensity", 
                      "23" = "Developed, Medium Intensity", "31" = "Barren Land", "41" = "Deciduous Forest", 
                      "42" = "Evergreen Forest", "43" = "Mixed Forest", "52" = "Shrub/Scrub", 
                      "72" = "Grassland/Herbaceous", "81" = "Pasture/Hay", "82" = "Crops", 
                      "90" = "Woody Wetlands", "95" = "Emergent Herbaceous Wetlands")

# Assuming landcover_raster_transformed is your raster object
raster_df <- as.data.frame(landcover_raster_transformed, xy = TRUE)
colnames(raster_df) <- c("x", "y", "landcover")
raster_df$landcover <- factor(raster_df$landcover, levels = landcover_values)

ggplot(raster_df) +
  # Plot the raster
  geom_raster(aes(x = x, y = y, fill = landcover)) +
  scale_fill_manual(values = landcover_colors, labels = landcover_labels, name = "Landcover Type") +
  
  # Overlay Alabama shape
  geom_sf(data = alabama_sf, fill = NA, color = "black", size = 0.5) +
  
  # Overlay Autauga WMA
  geom_sf(data = autauga_wma_sf, fill = NA, color = "red", size = 0.5) +
  
  # Define plot labels and limits
  ggtitle("Zoomed into Autauga WMA with Landcover") +
  coord_sf(xlim = c(wma_bbox$xmin, wma_bbox$xmax), ylim = c(wma_bbox$ymin, wma_bbox$ymax), expand = FALSE) +
  
  # Adjust plot theme
  theme_minimal() +
  theme(legend.position = "right")



```


# Add Autauga WMA Legend
```{r}

install.packages("ggspatial")
library(ggspatial)


ggplot(raster_df) +
  # Plot the raster
  geom_raster(aes(x = x, y = y, fill = landcover)) +
  scale_fill_manual(values = landcover_colors, labels = landcover_labels, name = "Landcover Type") +
  
  # Overlay Alabama shape
  geom_sf(data = alabama_sf, fill = NA, color = "black", size = 0.5) +
  
  # Overlay Autauga WMA with a named color for the legend
  geom_sf(data = autauga_wma_sf, fill = NA, color = "red", size = 0.5, aes(linetype = "Autauga WMA")) +
  
  # Set the name for the linetype legend
  scale_linetype_manual(values = "solid", labels = "Autauga WMA", name = "") +
  
  # Define plot labels and limits
  ggtitle("Landcover in Autauga WMA") +
  coord_sf(xlim = c(wma_bbox$xmin, wma_bbox$xmax), ylim = c(wma_bbox$ymin, wma_bbox$ymax), expand = FALSE) +
  
  # Customize the legend to combine landcover types and Autauga WMA boundary
  guides(fill = guide_legend(order = 1), 
         linetype = guide_legend(order = 2)) +
  
  # Adjust plot theme
  theme_minimal() +
  theme(legend.position = "right") +

   # Add north arrow
  annotation_north_arrow(location = "br", which_north = "true", pad_x = unit(0.5, "in"), pad_y = unit(0.5, "in")) +
  
    # Set custom labels for axes
  labs(x = "Longitude", y = "Latitude") +
  
  # Add scale bar
  annotation_scale(location = "br", width_hint = 0.5)


```

```{r}
ggplot() +
 # Overlay Alabama shape
  geom_sf(data = alabama_sf, fill = NA, color = "black", size = 0.5) +
  
  # Overlay Autauga WMA without a named color for the legend
  geom_sf(data = autauga_wma_sf, fill = NA, color = "red", size = 0.5) +
  
  # Add an extent rectangle for Autauga WMA with a label for the legend
  geom_rect(aes(xmin = wma_bbox$xmin, xmax = wma_bbox$xmax,
                ymin = wma_bbox$ymin, ymax = wma_bbox$ymax, linetype = "WMA Extent"),
            color = "blue", fill = NA, size = 0.5) +

    scale_linetype_manual(values = c("WMA Extent" = "dashed"), 
                        labels = c("WMA Extent"), 
                        name = "") +

    # Define plot labels
  ggtitle("Alabama with Autauga WMA Extent") +
  labs(x = "Longitude", y = "Latitude") +
  
  # Adjust plot theme
  theme_minimal() +
  theme(legend.position = "right",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))



```










# Add secondary alabama map

```{r}
library(ggplot2)
library(cowplot)

# Zoomed-in plot
zoomed_plot <- ggplot(raster_df) +
  geom_raster(aes(x = x, y = y, fill = landcover)) +
  scale_fill_manual(values = landcover_colors, labels = landcover_labels, name = "Landcover Type") +
  geom_sf(data = alabama_sf, fill = NA, color = "black", size = 0.5) +
  geom_sf(data = autauga_wma_sf, fill = NA, color = "red", size = 0.5) +
  coord_sf(xlim = c(wma_bbox$xmin, wma_bbox$xmax), ylim = c(wma_bbox$ymin, wma_bbox$ymax), expand = FALSE) +
  ggtitle("Zoomed into Autauga WMA with Landcover") +
  theme_minimal() +
  theme(legend.position = "right")


# Broader map with extent rectangle
broader_map <- ggplot() +
  geom_sf(data = alabama_sf, fill = NA, color = "black", size = 0.5) +
  geom_rect(aes(xmin = wma_bbox$xmin, xmax = wma_bbox$xmax,
                ymin = wma_bbox$ymin, ymax = wma_bbox$ymax),
            color = "blue", fill = NA, linetype = "dashed", size = 0.5) +
  ggtitle("Alabama with Autauga WMA Extent") +
  theme_minimal()

# Arrange the two plots with different widths
combined_plot <- plot_grid(zoomed_plot, broader_map, 
                           ncol = 2, 
                           rel_widths = c(3, 1), # Adjust the ratio as needed
                           align = 'v')

# Display the combined plot
print(combined_plot)



```


```{r}

 # for handling spatial data

# Load the shapefile
autauga_wma <- vect("AutaugaWMA/AutaugaWMA/AutaugaWMA.shp")

# Load Alabama shape
alabama <- states() %>% 
  filter(NAME == 'Alabama')

# Extract the CRS from the Alabama shape in WKT format
crs_alabama_wkt <- st_crs(alabama)$wkt

# Transform the CRS of Autauga WMA
autauga_wma_transformed <- terra::project(autauga_wma, crs_alabama_wkt)


```


```{r}
# Plot Alabama
plot(st_geometry(alabama), main = "Alabama with Autauga WMA")

# Overlay Autauga WMA
plot(autauga_wma_transformed, add = TRUE, col = "red")



```

```{r}
# Get the extent of autauga_wma_transformed
ext <- ext(autauga_wma_transformed)


# Plot Alabama
plot(st_geometry(alabama), main = "Alabama with Autauga WMA")

# Overlay Autauga WMA
plot(autauga_wma_transformed, add = TRUE, col = "red")


```

```{r}
# Get the extent of autauga_wma_transformed
ext <- ext(autauga_wma_transformed)

# Plot Alabama with zoom to autauga_wma_transformed
plot(st_geometry(alabama), main = "Zoomed Alabama with Autauga WMA", 
     xlim = c(ext[1], ext[2]), ylim = c(ext[3], ext[4]))

# Overlay Autauga WMA
plot(autauga_wma_transformed, add = TRUE, col = "red")

```

```{r}
# Get the extent of autauga_wma_transformed
ext <- ext(autauga_wma_transformed)

# Calculate buffer (adjust the buffer size as needed)
buffer_size <- 0.01  # Example buffer size, adjust as needed
buffered_ext <- c(ext[1] - buffer_size, ext[2] + buffer_size, ext[3] - buffer_size, ext[4] + buffer_size)

# Plot Alabama with buffered extent
plot(st_geometry(alabama), main = "Zoomed Alabama with Autauga WMA", 
     xlim = c(buffered_ext[1], buffered_ext[2]), ylim = c(buffered_ext[3], buffered_ext[4]))

# Overlay Autauga WMA
plot(autauga_wma_transformed, add = TRUE, col = "red")

# Add a rectangle around autauga_wma_transformed
rect(ext[1], ext[3], ext[2], ext[4], border = "blue", lwd = 2)


```





# Plot Autauga WMA Shapefile on top of Alabama

```{r}
alabama = states() %>% 
  filter(NAME=='Alabama')

plot(st_geometry(alabama))

st_crs(alabama)$proj4string

```


```{r}
# Transform the CRS of the landcover raster to match the Alabama shapefile
landcover_raster_transformed <- project(landcover_raster, crs(alabama))

# Plot landcover raster with specified extent
plot(landcover_raster_transformed, xlim = c(ext[1], ext[2]), ylim = c(ext[3], ext[4]), main = "Zoomed Alabama with Landcover and Autauga WMA")

# Overlay Alabama shape
plot(st_geometry(alabama), add = TRUE)

# Overlay Autauga WMA with transparent fill
plot(autauga_wma_transformed, add = TRUE, col = "#FF000000", border = "red")


```
```{r}
# Calculate buffer (adjust the buffer size as needed)
buffer_size <- 0.01  # Example buffer size, adjust as needed
buffered_ext <- c(ext[1] - buffer_size, ext[2] + buffer_size, ext[3] - buffer_size, ext[4] + buffer_size)

# Plot landcover raster with buffered extent
plot(landcover_raster_transformed, xlim = c(buffered_ext[1], buffered_ext[2]), ylim = c(buffered_ext[3], buffered_ext[4]), main = "Zoomed Alabama with Landcover and Autauga WMA")


# Overlay Alabama shape
plot(st_geometry(alabama), add = TRUE)

# Overlay Autauga WMA with transparent fill
plot(autauga_wma_transformed, add = TRUE, col = "#FF000000", border = "red")

# Add a rectangle around autauga_wma_transformed
rect(ext[1], ext[3], ext[2], ext[4], border = "blue", lwd = 2)


```

```{r}
# Plot landcover raster with buffered extent
plot(landcover_raster_transformed, xlim = c(buffered_ext[1], buffered_ext[2]), ylim = c(buffered_ext[3], buffered_ext[4]), main = "Zoomed Alabama with Landcover and Autauga WMA")

# Overlay Alabama shape
plot(st_geometry(alabama), add = TRUE)

# Overlay Autauga WMA with transparent fill
plot(autauga_wma_transformed, add = TRUE, col = "#FF000000", border = "red")

# Add a rectangle around autauga_wma_transformed
rect(ext[1], ext[3], ext[2], ext[4], border = "blue", lwd = 2)

# Add legend
legend("topright",  # Adjust position as needed
       legend = c("Alabama", "Autauga WMA", "Autauga WMA Extent"),
       fill = c(NA, "#FF000000", NA),
       border = c(NA, "red", "blue"),
       lwd = c(1, 2, 2))


```


```{r}

# Define color mapping
color_mapping <- setNames(landcover_colors, landcover_values)

# Now use this color mapping in your plot
plot(reclassed_raster, col = color_mapping[as.character(landcover_values)], 
     xlim = c(ext[1], ext[2]), ylim = c(ext[3], ext[4]), 
     main = "Zoomed Alabama with Landcover and Autauga WMA")

# The rest of your plotting code remains the same


# Landcover values
landcover_values <- c(11, 21, 22, 23, 31, 41, 42, 43, 52, 72, 81, 82, 90, 95)

# Corresponding colors
landcover_colors <- c("darkblue", "pink", "pink", "pink", "lightgray", "lightgreen", 
                      "darkgreen", "green", "lightyellow", "yellow", "orange", 
                      "brown", "lightblue", "purple")

# Landcover values and corresponding colors remain the same as before

# Reclassify the raster
reclassed_raster <- terra::classify(landcover_raster_transformed, 
                                    rcl = cbind(landcover_values, landcover_values))

# Apply the color palette
plot(reclassed_raster, col = color_mapping[as.character(landcover_values)], 
     xlim = c(ext[1], ext[2]), ylim = c(ext[3], ext[4]), 
     main = "Zoomed Alabama with Landcover and Autauga WMA", legend=F)

# Overlay Alabama shape
plot(st_geometry(alabama), add = TRUE)

# Overlay Autauga WMA with transparent fill
plot(autauga_wma_transformed, add = TRUE, col = "#FF000000", border = "red")

# Add a rectangle around autauga_wma_transformed
rect(ext[1], ext[3], ext[2], ext[4], border = "blue", lwd = 2)

# Landcover labels and custom legend
landcover_labels <- c("Open Water", "Developed", "Developed", "Developed", "Barren Land",
                      "Deciduous Forest", "Evergreen Forest", "Mixed Forest", "Shrub/Scrub",
                      "Grassland/Herbaceous", "Pasture/Hay", "Crops", "Woody Wetlands", 
                      "Emergent Herbaceous Wetlands")

# Add legend on the right side of the map
legend("right", legend = landcover_labels, fill = landcover_colors, cex = 0.7, bg = 'NA')

# Use par(xpd=TRUE) and specify coordinates for legend
par(xpd = TRUE)
legend(x = 5, y = 1, legend = landcover_labels, fill = landcover_colors, cex = 0.7, bg = 'NA', horiz = FALSE, xjust = 0)

```











# ggplot

```{r}
library(terra)
library(ggplot2)
library(dplyr)

# Convert raster to data frame
raster_df <- as.data.frame(reclassed_raster, xy = TRUE)
colnames(raster_df) <- c("x", "y", "landcover")

# Map landcover values to colors
raster_df$color <- color_mapping[as.character(raster_df$landcover)]

# Convert terra objects to sf
alabama_sf <- st_as_sf(alabama)
autauga_wma_sf <- st_as_sf(autauga_wma_transformed)

ggplot() +
  # Plot the raster
  geom_tile(data = raster_df, aes(x = x, y = y, fill = color), color = NA) +
  scale_fill_identity() +
  
  # Overlay Alabama shape
  geom_sf(data = alabama_sf, fill = NA, color = "black") +
  
  # Overlay Autauga WMA with transparent fill
  geom_sf(data = autauga_wma_sf, fill = "#FF000000", color = "red") +
  
  # Define plot labels
  ggtitle("Zoomed Alabama with Landcover and Autauga WMA") +
  
  # Define plot limits
  coord_sf(xlim = c(ext[1], ext[2]), ylim = c(ext[3], ext[4]), expand = FALSE) +
  
  # Add legend for landcover
  scale_color_manual(values = landcover_colors, 
                     labels = landcover_labels,
                     guide = guide_legend(override.aes = list(fill = landcover_colors)))


```

```{r}
# Transform the CRS of the landcover raster to match the Alabama shapefile
landcover_raster_transformed <- project(landcover_raster, crs(alabama))

# Convert raster to a data frame
raster_df <- as.data.frame(landcover_raster_transformed, xy = TRUE)
colnames(raster_df) <- c("x", "y", "value")

# Convert shapefiles to sf objects
alabama_sf <- st_as_sf(alabama)
autauga_wma_sf <- st_as_sf(autauga_wma_transformed)

ggplot() +
  # Plot the raster
  geom_raster(data = raster_df, aes(x = x, y = y, fill = value)) +
  scale_fill_viridis_c() +  # Adjust color scale as needed
  
  # Overlay Alabama shape
  geom_sf(data = alabama_sf, fill = NA, color = "black") +
  
  # Overlay Autauga WMA
  geom_sf(data = autauga_wma_sf, fill = NA, color = "red") +
  
  # Define plot labels
  ggtitle("Alabama with Landcover and Autauga WMA") +
  
  # Adjust plot theme
  theme_minimal()


```

```{r}
# Calculate Autuaga extent
wma_extent <- st_bbox(autauga_wma_sf)

ggplot() +
  # Plot the raster with new limits
  geom_raster(data = raster_df, aes(x = x, y = y, fill = value)) +
  scale_fill_viridis_c() +  # Adjust color scale as needed
  
  # Overlay Alabama shape
  geom_sf(data = alabama_sf, fill = NA, color = "black") +
  
  # Overlay Autauga WMA
  geom_sf(data = autauga_wma_sf, fill = NA, color = "red") +
  
  # Define plot labels
  ggtitle("Zoomed into Autauga WMA with Landcover") +
  
  # Set plot limits to the extent of the WMA
  coord_sf(xlim = c(wma_extent$xmin, wma_extent$xmax), 
           ylim = c(wma_extent$ymin, wma_extent$ymax), expand = FALSE) +
  
  # Adjust plot theme
  theme_minimal()


```

```{r}

# Landcover values and corresponding colors
landcover_colors <- c("11" = "blue",
                      "31" = "gray", "41" = "lightgreen", "42" = "darkgreen", 
                      "43" = "green", "52" = "yellow", "72" = "yellow", 
                      "81" = "red", "82" = "brown", "90" = "lightblue", "95" = "purple")

# Landcover labels
landcover_labels <- c("11" = "Open Water", "31" = "Barren Land", "41" = "Deciduous Forest", 
                      "42" = "Evergreen Forest", "43" = "Mixed Forest", "52" = "Shrub/Scrub", 
                      "72" = "Grassland/Herbaceous", "81" = "Pasture/Hay", "82" = "Crops", 
                      "90" = "Woody Wetlands", "95" = "Emergent Herbaceous Wetlands")


# Convert raster to a data frame
raster_df <- as.data.frame(reclassed_raster, xy = TRUE)
colnames(raster_df) <- c("x", "y", "landcover")

ggplot(raster_df) +
  # Plot the raster
  geom_raster(aes(x = x, y = y, fill = factor(landcover))) +
  scale_fill_manual(values = landcover_colors, labels = landcover_labels, name = "Landcover Type") +
  
  # Overlay Alabama shape
  geom_sf(data = alabama_sf, fill = NA, color = "black") +
  
  # Overlay Autauga WMA
  geom_sf(data = autauga_wma_sf, fill = NA, color = "red") +
  
  # Define plot labels and limits
  ggtitle("Zoomed Alabama with Landcover and Autauga WMA") +
  coord_sf(xlim = c(wma_extent$xmin, wma_extent$xmax), ylim = c(wma_extent$ymin, wma_extent$ymax), expand = FALSE) +
  
  # Adjust plot theme
  theme_minimal() +
  theme(legend.position = "right")


```


```{r}



# Updated corresponding colors (add 71 if needed)
landcover_colors <- c("11" = "darkblue", "21" = "red", "22" = "red", 
                      "31" = "black", "41" = "green", "42" = "green", 
                      "43" = "green", "52" = "red", "71" = "yellow",
                      "81" = "orange", "82" = "brown", "90" = "lightblue", 
                      "95" = "red") # Replace 'color_for_71' with the actual color

# Updated landcover labels (add 71 if needed)
landcover_labels <- c("11" = "Open Water", "21" = "Developed", "22" = "Developed", 
                       "31" = "Barren Land", "41" = "Deciduous Forest", 
                      "42" = "Evergreen Forest", "43" = "Mixed Forest", "52" = "Shrub/Scrub", "71" = "Grassland", 
                      "81" = "Pasture/Hay", "82" = "Crops", 
                      "90" = "Woody Wetlands", "95" = "Emergent Herbaceous Wetlands") # Add label for 71 if needed

ggplot(filtered_raster_df) +
  geom_raster(aes(x = x, y = y, fill = landcover)) +
  scale_fill_manual(values = landcover_colors, labels = landcover_labels, name = "Landcover Type") +
  
  # Overlay Alabama shape
  geom_sf(data = alabama_sf, fill = NA, color = "black", size = 0.5) +
  
  # Overlay Autauga WMA
  geom_sf(data = autauga_wma_sf, fill = NA, color = "red", size = 0.5) +
  
  # Define plot labels and limits
  ggtitle("Zoomed into Autauga WMA with Landcover") +
  coord_sf(xlim = c(wma_extent$xmin, wma_extent$xmax), ylim = c(wma_extent$ymin, wma_extent$ymax), expand = FALSE) +
  
  # Adjust plot theme
  theme_minimal() +
  theme(legend.position = "right", 
        legend.background = element_rect(fill = "white", colour = "black"))


```

# Load Alabama
```{r}
alabama = states() %>% 
  filter(NAME=='Alabama')

plot(st_geometry(alabama))

st_crs(alabama)$proj4string

```


# Plot WMAs on Alabama
```{r}

# Load the raster data
landcover_raster <- terra::rast("")

# Plot the landcover raster data
plot(landcover_raster)


```

# Plot WMAs on Alabama
```{r}

# Transform CRS of Alabama using sf
alabama_transformed <- st_transform(alabama, st_crs(wma_shapefile))

alabama_transformed_vect <- vect(alabama_transformed)

# Plot transformed Alabama
plot(alabama_transformed_vect, main = "Alabama")

# Plot WMA Shapefile in a new window
plot(wma_shapefile, main = "WMA Shapefile")

# Plot transformed Alabama
plot(alabama_transformed_vect, main = "Alabama with WMA Shapefile")

# Assuming wma_shapefile is of type lines or points
# Replace lines() with points() if it's a point dataset
lines(wma_shapefile, col = "red")

plot(alabama_transformed_vect, main = "Alabama with WMA Shapefile")
lapply(1:length(wma_shapefile), function(i) plot(wma_shapefile[i,], add=TRUE))



```



Let's get a first look at the spatial locations of our data. The locations were collected on a GPS unit using the NAD83 datum of the planet and the UTM zone 10 projection. So the x and y coordinates you see are UTMs.
```{r}
ggplot(siteData, aes(x=easting, y=northing))+
  geom_point()
```


Now we will turn the raw data frame into a spatial sf object by specifying which columns represent the coordinates and the datum/projection in which those coordinates are specified. This website is going to be helpful for you figuring out how to define your projection so that this function can understand it: https://spatialreference.org/.
```{r}
siteSf = st_as_sf(siteData, coords=c('easting', 'northing'), crs='+proj=utm +zone=10 +datum=NAD83 +units=m +no_defs +type=crs')

plot(st_geometry(siteSf))
```
Note that if you do not want to create a shapefile from scratch, you can read one in that you have already created using the st_read() function.

We are also going to need a spatial object that shows the extent of the area encompassed by our points so that we can plot it on our zoomed out map. To do that we first extract the x and y min and max for the bounding box (bbox) of the spatial points in our study area. Next we'll add 2 km to each dimension so that the bounding box is not pressed right up against our points. Finally, we'll convert it into a spatial object that we can plot.
```{r}

#Extract x and y coordinates of the bounding box
studyArea = st_bbox(siteSf)
studyArea

#Let's add 2000 m to each dimension
studyArea = studyArea + c(-2000, -2000, 2000, 2000)
studyArea

#And now we convert it to a spatial sfc object
studyArea = st_as_sfc(studyArea)

plot(studyArea)
```

Next we need a boundary file for the state of Oregon where our study took place. We are going to use that for the zoomed-out map where we show the context of our study region. Fortunately, it's easy to pull state shapefiles from the tigris package.

```{r}
oregon = states() %>% 
  filter(NAME=='Oregon')

plot(st_geometry(oregon))

st_crs(oregon)$proj4string
```
Note that the projection of the Oregon state shapefile is different than that of the GPS points we included earlier. That will have to be rectified to ensure we can plot these points in the same place. Here we are transforming the Oregon object to the same coordinate system as the siteSf object.
```{r}
oregon = st_transform(oregon, st_crs(siteSf))
```

Now we can overlay them all in the same place and see what we've got so far. This will form the basis for our zoomed-out map. We can see the state of Oregon, and an extent rectangle showing the area in which we worked. I've also added the points themselves for good measure.
```{r}
plot(st_geometry(oregon))
plot(studyArea, col='red', add=T)
plot(st_geometry(siteSf), add=T)
```
# Zoomed-in map

For our zoomed-in map, I'd like to add a couple of features that will help us visualize the study area. Namely, I'd like to demonstrate how mountainous the area is as well as the distribution of old forest in the region because Marbled Murrelets tend to prefer old forests. To start with, let's download some elevation data and crop it to the state of Oregon.

Note I am projecting the Oregon state boundary into the same CRS as the elevation data. In general, it is much faster to reproject shapefiles than raster files, especially large raster files.
```{r}
usaElev = elevation_30s(country='USA', path=tempdir())


oregon = st_transform(oregon, st_crs(usaElev))
oregonElev = crop(usaElev, oregon)

plot(oregonElev)
plot(st_geometry(oregon), add=T)
```
When I crop the data, the function crops the raster by the extent of the shape object (the bbox), but I would rather get rid of the values outside of the state boundaries. To do that, I can mask values that are not located inside of the shapefile. This just gives NA values to anything not located within the shape.

```{r}
oregonElev = mask(oregonElev, oregon)
plot(oregonElev)
```
But we aren't going to need elevation for the entire state, so let's zoom in on our study region.
```{r}
studyArea = st_transform(studyArea, st_crs(oregon))
studyAreaElevation = crop(oregonElev, studyArea)
plot(studyAreaElevation, col=sequential_hcl(13, "Terrain"))
```
Now I'm going to import a raster layer that shows the distribution of forests with an old-growth structural index (OGSI) of > 80, meaning that the trees in the forest have structural characteristics resembling trees > 80 years old. This file was created by the US Forest Service, and I downloaded it from a free online repository (https://lemma.forestry.oregonstate.edu/projects/nwfp). I have stored it on GitHub again, so it can be directly imported here.

```{r}
# ogsi80 = rast('https://github.com/ValenteJJ/SpatialEcology/blob/main/Week2/ogsi80_utm.tif?raw=true')
ogsi80 = rast('https://github.com/ValenteJJ/SpatialEcology/raw/main/Week2/ogsi80_utm.tif')
studyArea = st_transform(studyArea, st_crs(ogsi80))
ogsi80 = crop(ogsi80, studyArea)
ogsi80[is.na(ogsi80)] = 0
plot(ogsi80)
```
The white cells show you 30 m pixels where old growth forest has been identified. One of the downsides of working with high-resolution rasters, however, is that you can end up with a TON of cells, and thus a ton of data which can slow down or crash R. It can also take a very long time to render images with lots of cells. Because this is just for visualization purposes (and not for analysis), we are going to re-aggregate this raster at a coarser resolution to make it easier on ourselves.

```{r}
ogsi80Coarse = aggregate(ogsi80, fact=5, fun="mean")

plot(ogsi80Coarse)
```
Now we're seeing a new raster where cells have been re-aggregated to a resolution of 150 m. Cells were joined by calculating the average value of the 25 cells that were put together. I'm going to make it binary again by saying that any cell with < 50% old growth is going to be considered NOT old growth, and the others ARE old growth (again, just for display purposes).

```{r}
ogsi80Coarse[ogsi80Coarse < 0.5] = 0
ogsi80Coarse[ogsi80Coarse >= 0.5] = 1

plot(ogsi80Coarse)
```
Let's reclassify this raster so that instead of 0 and 1 values, we are identifying the cells as old growth or not old growth forest.


```{r}
tmp = data.frame('ID' = c(0, 1), 'category' = c(NA, 'Old growth'))
tmp
og = categories(ogsi80Coarse, value=tmp)

plot(og)
```



# Make a nice map

Now that we have created all of our rasters and shapefiles, we can put it together with ggplot to make a map. First, we need to make sure that all of our layers are in the same reference system. We bounced around a bit above, so let's take stock.
```{r}
#These are in lat/long
crs(studyAreaElevation, proj=T)
crs(oregon, proj=T)


#These are in utms
crs(siteSf, proj=T)
crs(og, proj=T)
st_crs(studyArea)$proj4string

```
I'm going to reproject everything to UTMs so that we can show a nice scale bar in kilometers. I am not sure we've mentioned this yet, but note that there are two different functions for transforming the crs of a vector (st_transform) and a raster (project).

```{r}
studyAreaElevation = project(studyAreaElevation, crs(siteSf, proj=T))
oregon = st_transform(oregon, crs(siteSf, proj=T))
```

Ok, we're ready to go. Let's start simply with our zoomed-out map. Note that ggplot has a built-in geom for plotting sf objects (geom_sf). If you load the tidyterra package, there is a geom_spatraster geom available as well. We'll use both here.


```{r}
zoomedOut = ggplot()+
  geom_sf(data = oregon)+
  geom_sf(data = studyArea, fill=NA, color='red')

zoomedOut
  
```
Now let's just clean this up a bit.

```{r}
zoomedOut = zoomedOut +
  theme_bw()+ #Gets rid of most of the colors
  theme(panel.grid=element_blank(), #Removes the gridlines
        axis.text=element_blank(), #Removes the axis text
        axis.ticks=element_blank(), #Removes the axis tick marks
        panel.border=element_blank()) #Removes the frame around the graph

zoomedOut
```
Now let's make our zoomed-in map. On this one, we are going to add elevation, old growth, and sampling point data. First I'll add on the elevation layer. Note that I'm making the background blue so it sort of looks like the ocean.

```{r}
zoomedIn = ggplot()+
  geom_spatraster(data = studyAreaElevation)+
  scale_fill_gradient(low='white', high='black', na.value='#99CCFF', labs(fill='Elevation'))

zoomedIn
```
When I add in the old growth forest layer, inserting a new_scale_fill() line is critical so it doesn't try to apply the previous fill declaration to the next raster.
```{r}
zoomedIn = zoomedIn + 
  new_scale_fill()+
  geom_spatraster(data=og, alpha=0.5)+
  scale_fill_manual(values=c('Old growth' = 'green'), na.value=NA, name='Old growth', labels='')

zoomedIn
```
Let's throw on the study area rectangle so it is clear what is being shown in the zoomed out map. Then we'll clean it up a bit.

```{r}
zoomedIn = zoomedIn+
  new_scale_fill()+
  theme_bw()+
  theme(panel.grid=element_blank(),
        axis.text.x=element_text(angle=90, vjust=0.5, size=6))

zoomedIn
  
```
And lets place our survey sites over top.

```{r}
zoomedIn = zoomedIn +
  geom_sf(data = siteSf, aes(pch=treatment), color='orange')+
  scale_shape_manual(values=c(19, 17), name='Site type')

zoomedIn
```
Ok, we now have a zoomed-in map and a zoomed-out map. Let's piece them together using one of my favorite packages, cowplot.

# Cowplotting these together

Cowplot offers tons of flexibility for sticking together different elements that are otherwise cumbersome to stick together. You can find lots of vignettes here (https://wilkelab.org/cowplot/articles/index.html). Let's try something very simple and just place these side-by-side.

```{r}
plot_grid(zoomedOut, zoomedIn)
```
Simple enough, but I'd like to see the zoomed-in map on one side, and then the zoomed-out map on the other side stacked on top of the legend. That might make things look a bit more compact. One neat thing you can do is pull the legend out of a ggplot figure and treat it as its own element.

```{r}
legend = get_legend(zoomedIn)
ggdraw(legend)
```
For my purposes, I actually think it might be nicer to have the legend oriented horizontally, so let's switch that up. Then I'm going to delete the legend from the zoomedIn object.

```{r}
legend = get_legend(zoomedIn+theme(legend.direction='horizontal'))
ggdraw(legend)

zoomedIn = zoomedIn + theme(legend.position='none')
```
So now I have a zoomed-in map, a zoomed-out map, and a legend. Let's first stick the zoomed-out map on top of the legend. Note that I am specifying that I want the zoomed-out map to be half the height of the legend.

```{r}
supportingFig = plot_grid(zoomedOut, ggdraw(legend), ncol=1, rel_heights = c(0.5, 1))

supportingFig
```
And finally, I can stick the zoomed-in map next to the supportingFig.

```{r}
plot_grid(zoomedIn, supportingFig, rel_widths=c(1, 0.5))
```
And export it to a jpg.
```{r}
ggsave2('finalMap.jpg', plot=plot_grid(zoomedIn, supportingFig, rel_widths=c(1, 1)), width=5, height=4, units='in', dpi=300)
```
# Things I hope you have taken away from this lab

1. How to convert a csv into a spatial object.
2. How to import data from GitHub.
3. Exposure to several valuable spatial R packages.
4. Creating extent boxes from other spatial files.
5. A few different approaches for plotting spatial data.
6. Transforming vectors and rasters between CRS'.
7. Basic tools for cropping, manipulating values, and re-scaling spatial objects.
8. Plotting spatial objects using ggplot.
9. A basic understanding of what Cowplot can do for you.